Drought frequency predicts plant life history strategies

### 2018-11-28 manuscript/manuscript.tex
edits from Kathryn

\documentclass[man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Satellite-detected droughts predict plant life history},
            pdfauthor={J. Grey Monroe, Brian Gill, Kathryn Turner, \& John K. McKay},
            pdfkeywords={drought adaptation, life history evolution, remote sensing,
phylogeography, herbaria records},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Satellite-detected droughts predict plant life history}
    \author{J. Grey Monroe\textsuperscript{1,2}, Brian Gill\textsuperscript{3},
Kathryn Turner\textsuperscript{4}, \& John K. McKay\textsuperscript{2}}
    \date{}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals (i.e.~monocarpic or semelparous)
which complete their seed to seed life cycle within a single calendar
year from perennials (i.e.~polycarpic or iteroparous) which can persist
over multiple years. Annual plants flower once, set seed, senesce, and
then die, spending at least some portion of the year as a seed, where
they are relatively protected from environmental stress. In contrast,
perennial plants can continue vegetative growth after reproduction and
must survive conditions experienced during all seasons. These represent
fundamentally different life history strategies, but the ecological
factors that explain their evolution and distributions remain
empirically uresolved (Friedman \& Rubin, 2015).\\
Classical theory predict shorter life spans in environments where adult
mortality is high (Charnov \& Schaffer, 1973; Franco \& Silvertown,
1996; Stearns, 1992). In plants, this has been extended to the
hypothessis that annuality is adaptive when it allows plants to escape
drought (Schaffer \& M, 1975). Lack of water is perhaps the greatest
threat to survival during vegetatitive or reproductive growth and
annuals can remain dormant (and protected as a seed) during drought.
Thus, environments with greater seasonal drought frequency may select
for annual life histories that complete reproduction prior to drought
prone seasons. Conversely, environments with less frequent drought may
select for perennial species, which may benefit from multiple bouts of
reproduction and competitive advantage by preventing recruitment of
annual species (Corbin \& D'Antonio, 2004). These predictions have been
supported by the association of annualilty with arid environments in
wild rice (Morishima, Sano, \& Oka, 1984) and \emph{Oenothera} (Evans,
Hearn, Hahn, Spangle, \& Venable, 2005). Additionally, annual and
perennial species of \emph{Nemesia} were qualitatively associated with
winter rather and summer rainfall environments respectively (Datson,
Murray, \& Steiner, 2008) and annual species of \emph{Scorzoneroides}
were associated with environments classified as unpredictable
(Cruz-Mazo, Buide, Samuel, \& Narbona, 2009). However, whether the
history frequency of drought events indeed predicts the distributions
annual or perennial life history strategies has yet to be tested in a
phylogenetic context with suitable sample sizes of distrubution and
climate data.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural collections to test these classic
hypotheses about the evolution of life history strategies within the
African endemic mustard genus, \emph{Heliophila} L. (Brassicaceae). If
annuallity is an adaptive strategy allowing plants to escape drought
prone seasons, then drought frequency should predict the distribution of
life history strategies across landscapes, and annual species should be
more commonly associated with drought prone regions than perennial
species. Furthermore, if annual species have adapted to escape drought
prone seasons as seeds, observations of annual species should be rare
during drought prone seasons. Phylogenetic relatedness can have
significant non-random effects on species distributions and life history
traits (Barrett, Harder, \& Worley, 1996), and therefore we assessed the
relationship between life history distribution and drought frequency in
a phylogenetically controlled background.

@@

All analyses were peformed using R. All data and the source code to
produce this manuscript are available at
\url{https://github.com/greymonroe/heliophila}.

@@

Remotely sensed data is a powerful tool for characterizing seasonal
patterns in drought because it is less limited in spatial and temporal
scope and resolution than weather stations or field observations
(AghaKouchak et al., 2015). To quantify the frequency of drought during
different seasons across landscapes, we used the remotely sensed
Vegetative Health Index (VHI), which measures landscape scale reductions
in plant cover and temperature conditions characteristic of drought
(Kogan, 2001). Generated from data collected by NOAA AVHRR satellites
since 1981, the VHI combines normalized difference vegetation index
(NDVI) derived measures of vegetative stress (Vegetative Condition Index
- VCI) with temperature stress indicated by anomalies in thermal spectra
(Temperature Condition Index - TCI). The VHI of year \(y\) during week
\(w\) of \([1,52]\) at pixel \(i\) is derived from the following
equations, where \(n\) is the number of years observed.

@@

where \(NDVI_{min} = min(NDVI_{1981,w,i}...NDVI_{1981+n,w,i})\) and
\(NDVI_{max} = max(NDVI_{1981,w,i}...NDVI_{1981+n,w,i})\)

@@

where \(T_{min} = min(T_{1981,w,i}...T_{1981+n,w,i})\) and
\(T_{max} = max(T_{1981,w,i}...T_{1981+n,w,i})\)

\[VHI_{y,w,i} = 0.5(VCI_{y,w,i}) + 0.5(TCI_{y,w,i})\] Thus, VHI
measurements are standardized according to conditions historically
observed at each locations. These measurements have been validated and
generally used for evaluating drought risk and predicting crop yields in
agriculture (e.g., Rojas, Vrieling, \& Rembold, 2011; Kogan et al.,
2016). But they also present a new tool to study seasonal patterns in
the frequency of drought across environments and to test hypotheses
about the effect of drought on ecological and evolutionary processes
(Kerr \& Ostrovsky, 2003). As such, the VHI has been applied recently to
study drought related ecology of natural species and proven useful for
predicting infraspecific variation in drought tolerance traits and genes
(Dittberner et al., 2018; Mojica et al., 2016; Monroe et al., 2018).
Here, we accessed VHI data at \(16km^2\) resolution from 1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species.

@@

\emph{Heliophila} is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions. These are among the most botanically diverse environments on
Earth and the estimated \textasciitilde{}50 \emph{Heliophila} species
are considered to be the most diverse genus of the family Brassicaceae
(Mummenhoff, Al-Shehbaz, Bakker, Linder, \& Mühlhausen, 2005). This
genus includes both perennial and annual species and this change in life
history strategy has likely arisen multiple independent times (Appel \&
Al-Shehbaz, 1997; Mummenhoff et al., 2005). Furthermore, the fine scale
climatic heterogeneity of Southern Africa is ideal for studying the
distribution of traits in relation to environmental parameters (Sayre et
al., 2013). We used life histories reported by Mummenhoff et al. (2005),
grouping species with annual or perennial life histories. Perenniality
was defined based any form of perennial life history (e.g., herbs,
shrubs, mixed, etc). We also repeated these analyses without \emph{H.
subulata} and \emph{H. minima} which have mixed life histories and found
similar results. Because the original authors didn't report the nature
of the mixed trait (Mummenhoff et al., 2005), (ie. plasticity
vs.~genetic variation), we classified these species here as perennial
since they have the capacity to maintain vegetative growth after
reproduction at least to some capacity.

@@

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years. Herbarium specimens and
their associated metadata have been used since the 1960s to study
species' geographical distributions (reviewed by Willis et al. (2017)
and Lang, Willems, Scheepens, Burbano, and Bossdorf (2018)). And as they
become digitized (Soltis, 2017), these collections have been used to
study relationships between trait distributions, geography, and climate
(Davis, Willis, Connolly, Kelly, \& Ellison, 2015; Stropp et al., 2016;
Václavı'k, Beckmann, Cord, \& Bindewald, 2017; Wolf, Zimmerman,
Anderegg, Busby, \& Christensen, 2016). To characterize the diributions
of annual and perennial \emph{Heliophila} species, all records for the
genus \emph{Heliophila} were downloaded from the Global Biodiversity
Information Facility (gbif.org) on July 21, 2018 (``GBIF occurrence
download,'' 2018).

@@

\begin{itemize}
\tightlist
\item
  Records for species with reported life history\\
\item
  Records with geospatial data\\
\item
  Records without known geospatial issues (i.e., coordinates reported
  are those of herbarium)\\
\item
  Records from collection sites classified as land pixels\\
\item
  Records from Africa\\
\item
  Records without duplicates (i.e., identical species, location,
  collection date)
\end{itemize}

@@

There were clear visual differences between the distributions of the 960
annual and the 1232 perennial \emph{Heliophila} observation records (see
Figure \ref{fig:speciesmaps} for maps of individual species). While
annual species were generally found in the western regions of South
Africa and the Namibia, primarily in the Cape Floristic Region and
Succulent Karoo (Figure \ref{fig:mapsboxplots}a), the occurances of
perennials extended to the east coast of South Africa (Figure
\ref{fig:mapsboxplots}b).

@@

Theory has predicted that annuality should be adaptive in places where
stresses such as drought are more common. Conversely, perenniality
should be adaptive in places where such stresses are less frequent. We
found that the frequency of drought was significantly higher at the
locations of occurance records for annual species. In terms of raw
observation records (Figure @\ref(fig:mapsboxplots)g), the frequency of
drought was signficantly higher at the location of annuals during the
winter (t = 10.65, p = 0.00), spring (t = 10.73, p = 0.00), summer (t =
12.67, p = 0.00), and fall (t = 15.26, p = 0.00). Because raw occurance
records do not account for variation in the number of records per
species (Table \ref{tab:speciesmeanstable})), we also tested whether
mean drought frequency values of each species were significantly
different between annuals and perennials by a Firths-penalized logistic
regression. We found that the mean drought frequencies were signficantly
higher (\(\alpha = 0.05\)) in annual species during the spring, summer,
and fall (Table \ref{tab:modelstable}, Figure \ref{fig:lineplots}a). We
further tested whether annual species are found in places where droughts
occur more frequently while controlling for the phylogentic relatedness
of \emph{Heliophila} species. This is important, because environmental
differences in species distributions can be confounded with demographic
history caused by acestry. We found that while controlling for
phylogeny, the mean drought frequencies were signficantly higher
(\(\alpha = 0.05\)) in annual species during the spring, summer, and
fall (Table \ref{tab:modelstable}, Figure \ref{fig:lineplots}a). These
findings indicate that common acestry alone does not explain differences
the drought frequencies experienced in the environemnts of annual and
perennial \emph{Heliophila}.

@@

\begin{figure}
\centering
\includegraphics{../figures/line_and_dates.pdf}
\caption{\label{fig:lineplots}Comparison (mean +- SE) of drought frequency across
seasons measured at the GBIF records of annual and perennial species of
Heliophila. (B) Collection dates of GBIF records of annual and perennial
species of Heliophila.}
\end{figure}

The preceding results indicated that annual speices are found in
environments where droughts are signifacantly more frequent, especially
in the summer and fall. Classic life history theory hypothesizes that
annuality reflects adaptation to such environments because it allows
species to escape stressful conditions. If this is the case, we would
expect that annuals spend the drought prone seasons of summer and fall
as seeds. To test this hypothesis, we compared the dates of occurance
records between annual and perennial \emph{Heliophila} species. The
distributions reveal a considerable difference in the timing of
observation of these two life histories. In comparison to perennials,
which appear to be collected throughout the year, annuals are almost
exclusively observed during the spring (Figure \ref{fig:lineplots}b).
The differences between the distrubution of collection dates were
signicant by all tests (ks.test D = 0.25, p = 0; bartlett.test K2 =
503.18, p = 0.00) This is consistent with a model of life history in
which annual species flower in the spring, set seed, senesce, and die
before the summer. Thus, these annual species are likely to remain
dormant during the summer and fall, when drought is the strongest
predictor of the distrubutions of annual and perennial life histories
(Figure \ref{fig:lineplots}a).

@@

To test the hypothesis that annual and perennial plants reflect
adaptation to alternative drought environments we examined the landscape
distribution of life history strategies in the large and diverse mustard
genus, \emph{Heliophila}. Using metadata of 2192 occurance records and a
34 year dataset of satellite-detected droughts, we tested the prediction
that annual species are more often observed in drought-prone locations
than perennial species, when controlling for phylogenetic relatedness.
We found that drought frequency is significantly different between the
distributions of annual and perennial species, with annuals being found
in environments with more frequent drought, and that this signal is
strongest during the summer. These results remain significant while
controlling for the phylogenetic relationships of \emph{Heliophila}
species, yielding support for the role that natural selection has played
in driving contemporary distributions of these alternatives strategies
in relation to drought regimes.

@@

These findings support classical theoretical predictions about the
adaptive value of annual and perennial life history strategies. Taken
together, they suggest that in \emph{Heliophila}, annual species are
adapted to environments with increased summer droughts by avoiding these
seasons in a dormant seed phase of their life cycle. They also suggest
that perenniality is adaptive in environemnts where droughts are less
frequent. While most previous work has focused on describing the
evolutionary origins of annuality (Andreasen \& Baldwin, 2001; Barrett
et al., 1996; Conti, Soltis, Hardig, \& Schneider, 1999; Friedman \&
Rubin, 2015; Verboom, Linder, \& Stock, 2004) there are at least a few
other cases where perenniality appears to have arisen from an annual
acestor (Bena, Lejeune, Prosperi, \& Olivieri, 1998; Tank \& Olmstead,
2008). And while early theory predicted selection for annuality when
adult morality is high (Stearns, 1992), we also find evidence that the
transition to perenniality could be explained by historical drought
regimes. The phylogeny reveals several transitions from annual to
perennial life history (Figure \ref{fig:phylogeny}a) and that perennials
distributions extend into regions where drought frequency is low (Figure
\ref{fig:mapsboxplots}b,Figure \ref{fig:speciesmaps}). Perennials may be
able to out complete annual relatives in environments where the
infrequency of drought favors strategies that allow plants to benefit
from growth over many seasons. It may also indicate that annuals rely on
drought as a source of disturbance for seedling recruitment when
competing with perennials (Corbin \& D'Antonio, 2004). Indeed, no annual
species were observed in the low drought regions of east South Africa
\ref{fig:mapsboxplots}a,Figure \ref{fig:speciesmaps}).

These findings suggest that species with locally adapted life history
strategies could be threatened by rapidly changing drought regimes (Dai,
2011). This could have impacts on ecosystem functioning if the
composition of annual and perennial species changes as a response
(Garnier, Cordonnier, Guillerm, \& Sonié, 1997; Roumet, Urcelay, \&
Dı'az, 2006). Furthermore, the frequency of drought may be an important
factor when considering the use of perennial cropping sysyems (Lelièvre
\& Volaire, 2009; Parry, Flexas, \& Medrano, 2005).

In conclusion, we find strong support for classic life history theory
which predicts that annuality is adaptive environments where droughts
occur more frequently. Additionally, we report evidence consistent with
a life history model in annuals in which they escape drought prone
seasons during the seed phase of their life cycle. Finally, we find
evidence that the distributions of perennial lineages may indicate a
competitive advantage in areas where droughts are infrequent. More
broadly, this work demonstrates the power of emerging data to address
outstanding classic hypotheses in ecology and evolution.

@@

We thank \_\_ for generous feedback that improved the quality of this
work. This research was supported by NSF Award 1701918 and USDA-NIFA
Award 2014-38420-21801 to JGM.

>>>>

\documentclass[man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Drought frequency predicts plant life history strategies},
            pdfauthor={J. Grey Monroe, Brian Gill, Kathryn G. Turner, \& John K. McKay},
            pdfkeywords={drought adaptation, life history evolution, remote sensing,
phylogeography, herbaria records},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Drought frequency predicts plant life history strategies}
    \author{J. Grey Monroe\textsuperscript{1,2}, Brian Gill\textsuperscript{3},
Kathryn G. Turner\textsuperscript{4}, \& John K.
McKay\textsuperscript{2}}
    \date{}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals which complete their seed to
seed life cycle within a single calendar year from perennials which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed, where they are relatively protected from environmental stress. In
contrast, perennial plants can continue vegetative growth after
reproduction and must survive conditions experienced during all seasons.
These represent fundamentally different life history strategies, but the
ecological factors that explain their evolution and distributions remain
empirically uresolved (Friedman \& Rubin, 2015).\\
Classical theory predict shorter life spans in environments where adult
mortality is high (Charnov \& Schaffer, 1973; Franco \& Silvertown,
1996; Stearns, 1992). In plants, this has been extended to the
hypothessis that annuality is adaptive when it allows plants to escape
drought (Schaffer \& Gadgil, 1975). Lack of water is perhaps the
greatest threat to survival during vegetatitive or reproductive growth
and annuals can remain dormant (and protected as a seed) during drought.
Thus, environments with greater seasonal drought frequency may select
for annual life histories that complete reproduction prior to drought
prone seasons. Conversely, environments with less frequent drought may
select for perennial species, which may benefit from multiple bouts of
reproduction and competitive advantage by preventing recruitment of
annual species (Corbin \& D'Antonio, 2004). These predictions have been
supported by the association of annualilty with arid environments in
\emph{Oryza perennis} (Morishima, Sano, \& Oka, 1984) and
\emph{Oenothera} (Evans, Hearn, Hahn, Spangle, \& Venable, 2005).
Additionally, annual and perennial species of \emph{Nemesia} were
qualitatively associated with winter rather and summer rainfall
environments respectively (Datson, Murray, \& Steiner, 2008) and annual
species of \emph{Scorzoneroides} were associated with environments
classified as unpredictable (Cruz-Mazo, Buide, Samuel, \& Narbona,
2009). However, whether the history frequency of drought events indeed
predicts the distributions annual or perennial life history strategies
has yet to be tested in a phylogenetic context with suitable sample
sizes of distrubution and climate data.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses about the evolution of life history strategies within
the African endemic mustard genus, \emph{Heliophila} L. (Brassicaceae).
If annuallity is an adaptive strategy allowing plants to escape drought
prone seasons, then drought frequency should predict the distribution of
life history strategies across landscapes, and annual species should be
more commonly associated with drought prone regions than perennial
species. Furthermore, if annual species have adapted to escape drought
prone seasons, observations of growing annual species (i.e.~occurring in
forms other than seed) should be rare during drought prone seasons.
Phylogenetic relatedness can have significant non-random effects on
species distributions and life history traits (Barrett, Harder, \&
Worley, 1996), and therefore we assessed the relationship between life
history distribution and drought frequency in a phylogenetically
controlled background.

@@

All analyses were peformed using R. All data and the source code to
produce this manuscript are available at
\url{https://github.com/greymonroe/heliophila}. Softare used is listed
in the supplement.

@@

Remotely sensed data is a powerful tool for characterizing seasonal
patterns in drought because it is less limited in spatial and temporal
scope and resolution than weather stations or field observations
(AghaKouchak et al., 2015). To quantify the frequency of drought during
different seasons across landscapes, we used the remotely sensed
Vegetative Health Index (VHI), which measures landscape scale reductions
in plant cover and temperature conditions characteristic of drought
(Kogan, 2001). Generated from data collected by NOAA AVHRR satellites
since 1981, the VHI combines Normalized Difference Vegetation Index
(NDVI) derived measures of vegetative stress (Vegetative Condition Index
- VCI) with temperature stress indicated by anomalies in thermal spectra
(Temperature Condition Index - TCI). The VHI of year \(y\) during week
\(w\) of \([1,52]\) at pixel \(i\) is derived from the following
equations, where \(n\) is the number of years observed.

@@

@@

\[VHI_{y,w,i} = 0.5(VCI_{y,w,i}) + 0.5(TCI_{y,w,i})\] where
\(NDVI_{min} = min(NDVI_{1981,w,i}...NDVI_{1981+n,w,i})\) and
\(NDVI_{max} = max(NDVI_{1981,w,i}...NDVI_{1981+n,w,i})\) and
\(T_{min} = min(T_{1981,w,i}...T_{1981+n,w,i})\) and
\(T_{max} = max(T_{1981,w,i}...T_{1981+n,w,i})\)

Thus, VHI measurements are standardized according to conditions
historically observed at each locations. These measurements have been
validated and generally used for evaluating drought risk and predicting
crop yields in agriculture (e.g., Rojas, Vrieling, \& Rembold, 2011;
Kogan et al., 2016). But they also present a new tool to study seasonal
patterns in the frequency of drought across environments and to test
hypotheses about the effect of drought on ecological and evolutionary
processes (Kerr \& Ostrovsky, 2003). As such, the VHI has been applied
recently to study drought related ecology of natural species and proven
useful for predicting infraspecific variation in drought tolerance
traits and genes (Dittberner et al., 2018; Mojica et al., 2016; Monroe
et al., 2018). Here, we accessed VHI data at \(16km^2\) resolution from
1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species.

@@

\emph{Heliophila} is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions. These are among the most botanically diverse environments on
Earth and the estimated \textasciitilde{}50 \emph{Heliophila} species
occurring there are considered to make up the most diverse genus of the
family Brassicaceae (Mandáková et al., 2012; Mummenhoff, Al-Shehbaz,
Bakker, Linder, \& Mühlhausen, 2005). This genus includes both perennial
and annual species and this change in life history strategy has likely
arisen multiple independent times (Appel \& Al-Shehbaz, 1997; Mummenhoff
et al., 2005). Furthermore, the fine scale climatic heterogeneity of
Southern Africa is ideal for studying the distribution of traits in
relation to environmental parameters (Sayre et al., 2013). We used life
histories reported by Mummenhoff et al. (2005), grouping species with
annual or perennial life histories. Perenniality was defined based any
form of perennial life history (e.g., herbs, shrubs, mixed, etc). We
also repeated these analyses without \emph{H. subulata} and \emph{H.
minima} which have mixed life histories and found similar results.
Because the original authors didn't report the nature of the mixed trait
(Mummenhoff et al., 2005), (ie. plasticity vs.~genetic variation), we
classified these species here as perennial since they have the capacity
to maintain vegetative growth after reproduction at least to some
capacity.

@@

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years. Herbarium specimens and
their associated metadata have been used since the 1960s to study
species' geographical distributions (reviewed by Willis et al. (2017)
and Lang, Willems, Scheepens, Burbano, and Bossdorf (2018)). And as they
become digitized (Soltis, 2017), these collections have been used to
study relationships between trait distributions, geography, and climate
(Davis, Willis, Connolly, Kelly, \& Ellison, 2015; Stropp et al., 2016;
Václavı'k, Beckmann, Cord, \& Bindewald, 2017; Wolf, Zimmerman,
Anderegg, Busby, \& Christensen, 2016). To characterize the diributions
of annual and perennial \emph{Heliophila} species, all records for the
genus \emph{Heliophila} were downloaded from the Global Biodiversity
Information Facility (gbif.org) on July 21, 2018 (GBIF, 2018).

@@

\begin{itemize}
\tightlist
\item
  records for species with reported life history\\
\item
  records with geospatial data\\
\item
  records without known geospatial coodinate issues (i.e., coordinates
  reported are those of herbarium)\\
\item
  records from collection sites classified as land pixels\\
\item
  records from Africa (to exclude locations of cultivation)
\item
  records without duplicates (i.e., identical species, location,
  collection date)
\end{itemize}

@@

There were clear visual differences between the distributions of the 960
annual and the 1232 perennial \emph{Heliophila} observation records (see
Figure \ref{fig:speciesmaps} for maps of individual species). While
annual species were generally found in the western regions of South
Africa and Namibia, primarily in the Cape Floristic Region and Succulent
Karoo (Figure \ref{fig:mapsboxplots}a), the occurances of perennials
extended to the east coast of South Africa (Figure
\ref{fig:mapsboxplots}b).

@@

Theory predicts that annuality should be adaptive in places where
stresses such as drought are more common. Conversely, perenniality
should be adaptive in places where such stresses are less frequent. We
found that the frequency of drought was significantly higher at the
locations of occurance records for annual species. In terms of raw
observation records (Figure \ref{fig:mapsboxplots}g), the frequency of
drought was signficantly higher at the location of annuals during the
winter (t = 10.65, p = 0.00), spring (t = 10.73, p = 0.00), summer (t =
12.67, p = 0.00), and fall (t = 15.26, p = 0.00). Because raw occurance
records do not account for variation in the number of records per
species (Table \ref{tab:speciesmeanstable}), we also tested whether mean
drought frequency values of each species were significantly different
between annuals and perennials by a Firths-penalized logistic
regression. We found that the mean drought frequencies were signficantly
higher (\(\alpha = 0.05\)) in annual species during the spring, summer,
and fall (Table \ref{tab:modelstable}, Figure \ref{fig:lineplots}a). We
further tested whether annual species are found in places where droughts
occur more frequently while controlling for the phylogentic relatedness
of \emph{Heliophila} species. This is important, because environmental
differences in species distributions can be confounded with demographic
history caused by acestry. We found that while controlling for
phylogeny, the mean drought frequencies were signficantly higher
(\(\alpha = 0.05\)) in annual species during the spring, summer, and
fall (Table \ref{tab:modelstable}, Figure \ref{fig:lineplots}a). These
findings indicate that common acestry alone does not explain differences
the drought frequencies experienced in the environemnts of annual and
perennial \emph{Heliophila}.

@@

\begin{figure}
\centering
\includegraphics{../figures/line_and_dates.pdf}
\caption{\label{fig:lineplots}(a) Comparison (mean +- SE) of drought frequency across
seasons measured at the GBIF records of annual and perennial species of
Heliophila. (b) Collection dates of GBIF records of annual and perennial
species of Heliophila.}
\end{figure}

The preceding results indicated that annual speices are found in
environments where droughts are signifacantly more frequent, especially
in the summer and fall. Classic life history theory hypothesizes that
annuality reflects adaptation to such environments because it allows
species to escape stressful conditions. If this is the case, we would
expect that annuals spend the drought prone seasons of summer and fall
as seeds. To test this hypothesis, we compared the dates of occurance
records between annual and perennial \emph{Heliophila} species. The
distributions reveal a considerable difference in the timing of
observation of these two life histories. In comparison to perennials,
which appear to be collected throughout the year, annuals are almost
exclusively observed during the winter and spring (Figure
\ref{fig:lineplots}b). The differences between the distrubution of
collection dates were signicant by all tests (ks.test D = 0.25, p = 0;
bartlett.test K2 = 503.18, p = 0.00) This is consistent with a model of
life history in which annual species flower in the spring, set seed,
senesce, and die before the summer. Thus, these annual species are
likely to remain dormant during the summer and fall, when drought is the
strongest predictor of the distrubutions of annual and perennial life
histories (Figure \ref{fig:lineplots}a).

@@

To test the hypothesis that annual and perennial plants reflect
adaptation to alternative drought environments we examined the landscape
distribution of life history strategies in the large and diverse mustard
genus, \emph{Heliophila}. Using metadata of 2192 occurance records and a
34 year dataset of satellite-detected droughts, we tested the prediction
that annual species are more often observed in drought-prone locations
than perennial species, when controlling for phylogenetic relatedness.
We found that drought frequency is significantly different between the
distributions of annual and perennial species, with annuals being found
in environments with more frequent drought, and that this signal is
strongest during the seasons when annuals are likely escaping via seed
dormancy These results remain significant while controlling for the
phylogenetic relationships of \emph{Heliophila} species, yielding
support for the role that natural selection has played in driving
contemporary distributions of these alternatives strategies in relation
to drought regimens.

@@

These findings support classical theoretical predictions about the
adaptive value of annual and perennial life history strategies. Taken
together, they suggest that in \emph{Heliophila}, annual species are
adapted to environments with increased summer droughts by avoiding these
seasons in a dormant seed phase of their life cycle. They also suggest
that perenniality is adaptive in environemnts where droughts are less
frequent. While most previous work has focused on describing the
evolutionary origins of annuality (Andreasen \& Baldwin, 2001; Barrett
et al., 1996; Conti, Soltis, Hardig, \& Schneider, 1999; Friedman \&
Rubin, 2015; Verboom, Linder, \& Stock, 2004) there are at least a few
other cases where perenniality appears to have arisen from an annual
acestor (Bena, Lejeune, Prosperi, \& Olivieri, 1998; Tank \& Olmstead,
2008). And while early theory predicted selection for annuality when
adult morality is high (Stearns, 1992), we also find evidence that the
transition to perenniality could be explained by historical drought
regimens. The phylogeny reveals several transitions from annual to
perennial life history (Figure \ref{fig:phylogeny}a) and that the
distributions of perennial \emph{Heliophila} extend into regions where
drought frequency is low (Figure \ref{fig:mapsboxplots}b,Figure
\ref{fig:speciesmaps}). Perennials may be able to out complete annual
relatives in environments where the infrequency of drought favors
strategies that allow plants to benefit from growth over many seasons.
It may also indicate that annuals rely on drought as a source of
disturbance for seedling recruitment when competing with perennials
(Corbin \& D'Antonio, 2004). Indeed, no annual species were observed in
the low drought regions of eastern South Africa
\ref{fig:mapsboxplots}a,Figure \ref{fig:speciesmaps}).

These findings suggest that species with locally adaptive life history
strategies could be threatened by rapidly changing drought regimens
(Dai, 2011). This could have impacts on ecosystem functioning and
processes such as carbon cycling if the composition of annual and
perennial species changes as a response {[}@garnier1997specific; Roumet,
Urcelay, and Dı'az (2006);monroe2018ecoevolutionary{]}. Furthermore, the
frequency of drought may be an important factor when considering the use
of perennial cropping sysyems (Lelièvre \& Volaire, 2009; Parry, Flexas,
\& Medrano, 2005).

In conclusion, we find strong support for classic life history theory
which predicts that annuality is adaptive in environments where droughts
occur more frequently. Additionally, we report evidence consistent with
a life history model in annuals in which they escape drought prone
seasons during the seed phase of their life cycle. Finally, we find
evidence that the distributions of perennial lineages may indicate a
competitive advantage in areas where droughts are infrequent. More
broadly, this work highlights the irreplaceable value of natural history
collections and demonstrates the power of combining such information
with large scale remote sensing data to address outstanding classic
hypotheses in ecology and evolution.

@@

We thank Jesse Lasky and members of the Sloan lab for generous feedback
that improved the quality of this work. This research was supported by
NSF Award 1701918 and USDA-NIFA Award 2014-38420-21801 to JGM.
